Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-03-10 paper/paper1/ms.tex
Add a flowchart schematic, fill in placeholders, and fill in the table of symbols

\documentclass[modern]{aastex631}
\usepackage{hyperref}
\bibliographystyle{aasjournal}
\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}

@@

Existing open-source frameworks have overcome some of these challenges, or have been purpose-built for specialized applications.
These frameworks include \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, \texttt{specmatch} \citep{2015PhDT........82P}, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y}, \texttt{wobble} \citep{2019AJ....158..164B}, \texttt{starfish} \citep{czekala15}, \texttt{sick} \citep{2016ApJS..223....8C}, \texttt{psoap} \citep{2017ApJ...840...49C}, \texttt{FAL} \authorcomment1{Cargile et al. in prep}, CHIMERA \citep{2015ApJ...807..183L}, the \texttt{Cannon} \citep{2017ApJ...836....5H},  \texttt{MINESweeper} \citep{2020ApJ...900...28C}, and recently \texttt{ExoJAX} \citep{2022ApJS..258...31K}.
The design of these frameworks necessarily have to make a choice in the bias-variance tradeoff: is the tool more \emph{data}-driven or more \emph{model}-driven?  The statistical tradeoff can be viewed as a concession in physical self-consistency for model flexibility: more or fewer parameters; more \emph{accurate} or more \emph{precise}.

@@

\subsection{Voigt profile evaluation}
The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed-form using the Fadeeva function\authorcomment1{citation}. Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable. Here we adopt the pseudo-Voigt approximation \citep{Ida:nt0146}.

@@

@@

The optimizer updates the $a_j, \sigma_j, \gamma_j$ parameters by a small fraction of the Jacobian---called the learning rate LR---towards the direction that would improve the fit, for all $N_{\mathrm{lines}} \times 3$ parameters simultaneously. The Jacobian is calculated behind the scenes with automatic differentiation\authorcomment1{Citation} implemented as the so-called backpropagation algorithm or simply ``backprop''. We choose the \texttt{PyTorch} framework that computes these Jacobians efficiently for all of the mathematical primitives in our \texttt{blas\'e} implementation.

In principle, we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch. Further, an autodiff-aware implementation of the \emph{celerit\'e} algorithm\authorcomment1{add DFM citation} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting. In practice, PyTorch also offers hardware acceleration (on GPU or TPU)\authorcomment1{citation needed}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks. The GPyTorch framework\authorcomment1{add citation} may offer a workaround to the absence of a \emph{celerit\'e} implementation. Still, the performance of an analytic Jacobian implementation over the autodiff route presented here offers an interesting avenue of future research.

@@

Minibatches offer a natural solution to the line-by-line computational bottleneck while also providing optimization performance improvements. Minibatches act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent\authorcomment1{citation}. The Machine Learning/Artificial Intelligence (ML/AI) community employs minibatches to handle similar situations, in which the dataset volume exceeds the available GPU RAM. Our configuration departs slightly from the typical ML/AI situation since their models are usually ``small'' relative to their datasets. Here our model is ``big'' relative to our dataset.

@@

% Table
\begin{deluxetable}{cc}
    \tablecaption{Notation used in this paper\label{table2}}
    \tablehead{
        \colhead{Symbol} & \nocolhead{Meaning}
    }
    \startdata
    $N_{pix}$ & Number of pixels in the precomputed spectrum \\
    $N_{lines}$ & Number of spectral lines \\
    $\sigma_G$ & Gaussian line profile scale and standard deviation \\
    $\gamma_L$ & Lorentzian line profile half width\\
    $\pm \Delta \lambda_{\mathrm{buffer}}$ & Buffer exceeding the red and blue limits of data spectrum
    \enddata
\end{deluxetable}

>>>>

\documentclass[modern]{aastex631}
\usepackage{hyperref}
\bibliographystyle{aasjournal}
%\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}

@@

Existing open-source frameworks have overcome some of these challenges, or have been purpose-built for specialized applications.
These frameworks include \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, \texttt{specmatch} \citep{2015PhDT........82P}, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y}, \texttt{wobble} \citep{2019AJ....158..164B}, \texttt{starfish} \citep{czekala15}, \texttt{sick} \citep{2016ApJS..223....8C}, \texttt{psoap} \citep{2017ApJ...840...49C}, \texttt{FAL} (Cargile et al. \emph{in prep}), CHIMERA \citep{2015ApJ...807..183L}, the \texttt{Cannon} \citep{2017ApJ...836....5H},  \texttt{MINESweeper} \citep{2020ApJ...900...28C}, and recently \texttt{ExoJAX} \citep{2022ApJS..258...31K}.
The design of these frameworks necessarily have to make a choice in the bias-variance tradeoff: is the tool more \emph{data}-driven or more \emph{model}-driven?  The statistical tradeoff can be viewed as a concession in physical self-consistency for model flexibility: more or fewer parameters; more \emph{accurate} or more \emph{precise}.

@@

\subsection{Voigt profile evaluation}
The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed-form using the Voigt-Hjerting function \citep{1938ApJ....88..508H} as the real part of the complex Fadeeva function \citep[\emph{e.g.}][]{2011arXiv1106.0151Z}. Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable. Here we adopt the pseudo-Voigt approximation \citep{Ida:nt0146}.

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.98\textwidth]{figures/blase_flowchart01.png}
    \caption{Visual flowchart of \texttt{blas\'e}.  The technique starts with a precomputed synthetic stellar spectrum and a precomputed synthetic telluric spectrum with properties close to those previously estimated for target.  Both the stellar and telluric spectra get cloned.  The stellar model is warped to its extrinsic properties, and then the stellar and telluric models get multiplied together.  Next this joint model is convolved with an instrumental kernel and resampled to the data spectrum.  Finally the joint model parameters are optimized to match the data, subject to regularization (not shown).  The final products are semi-empirical super-resolution stellar and telluric templates, the $RV$ and $v\sin{i}$ of the star, and the line amplitude, location, width, and shape of all 10,000+ stellar and telluric lines. \authorcomment1{Some of the spectra are placeholders until we re-run the end-to-end-fitting.  Feedback on this figure layout is sought before investing fine-adjusting the spectra cutouts.} }
    \label{blase_flowchart}
\end{figure}

@@

The optimizer updates the $a_j, \sigma_j, \gamma_j$ parameters by a small fraction of the Jacobian---called the learning rate LR---towards the direction that would improve the fit, for all $N_{\mathrm{lines}} \times 3$ parameters simultaneously. The Jacobian is calculated behind the scenes with automatic differentiation implemented as the so-called backpropagation algorithm or simply ``backprop'' \citep{2015arXiv150205767G}. We choose the \texttt{PyTorch} framework that computes these Jacobians efficiently for all of the mathematical primitives in our \texttt{blas\'e} implementation.

In principle, we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch. Further, an autodiff-aware implementation of the \emph{celerit\'e} algorithm \citep{2017AJ....154..220F} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting. In practice, PyTorch also offers hardware acceleration on GPU or TPU \citep{10.1145/3079856.3080246}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks. The GPyTorch framework \citep{2018arXiv180911165G} may offer a workaround to the absence of a \emph{celerit\'e} implementation. Still, the performance of an analytic Jacobian implementation over the autodiff route presented here offers an interesting avenue of future research.

@@

Minibatches offer a natural solution to the line-by-line computational bottleneck while also providing optimization performance improvements. Minibatches act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent \citep{2016arXiv160904747R}. The Machine Learning/Artificial Intelligence (ML/AI) community employs minibatches to handle similar situations, in which the dataset volume exceeds the available GPU RAM. Our configuration departs slightly from the typical ML/AI situation since their models are usually ``small'' relative to their datasets. Here our model is ``big'' relative to our dataset.

@@

% Table
\begin{deluxetable}{cl}
    \tablecaption{Notation used in this paper\label{table2}}
    \tablehead{
        \colhead{Symbol} & \colhead{Meaning}
    }
    \startdata
    \hline
    \multicolumn{2}{c}{Spectra}\\
    \hline
    $\bm{\lambda}_S$ & Native wavelength coordinates of the precomputed stellar spectrum\\
    $\bm{\lambda}_T$ & Native wavelength coordinates of the telluric spectrum\\
    $\bm{\lambda}_D$ & Native wavelength coordinates of the data spectrum\\
    $\mathsf{S}$ & Flux values of the precomputed synthetic stellar spectral model $\bm{\lambda}_S$\\
    $\mathsf{T}$ & Transmission values of the precomputed synthetic telluric model \\
    $\mathsf{D}$ & The observed data spectrum flux values\\
    $\mathsf{B}$ & Blackbody of temperature $T_{\mathrm{eff}}$ to coarsely normalize $\mathsf{S}$\\
    $\mathsf{P}$ & Smooth polynomial to refine continuum-normalization of $\mathsf{S/B}$\\
    $\mathsf{S}_{\rm clone}$ & Evaluable and tunable cloned flux model of $\mathsf{S}$\\
    $\mathsf{T}_{\rm clone}$ & Evaluable and tunable cloned transmission model of $\mathsf{T}$\\
    $\mathsf{T}_{\rm resamp}$ & The cloned telluric model resampled from $\bm{\lambda}_T$ to $\bm{\lambda}_S$\\
    $\mathsf{S}_{\rm ext}$ & An augmentation of $\mathsf{S}_{\rm clone}$ with $v\sin{i}$ convolution and $RV$ translation\\
    $\mathsf{M}_{\rm joint}$ & The joint stellar and telluric model: $\mathsf{S}_{\rm ext} \cdot \mathsf{T}_{\rm resamp}$  \\
    $\mathsf{M}$ & Joint model convolved with instrument resolving power and resampled to $\bm{\lambda}_D$\\
    \hline
    \multicolumn{2}{c}{Line properties}\\
    \hline
    $\lambda_{c,i}$ & Line center position of the $i^{th}$ spectral line\\
    $a_i$ & Gaussian line profile amplitude of the $i^{th}$ spectral line \\
    $\sigma_i$ & Gaussian line profile scale of the $i^{th}$ spectral line\\
    $\gamma_i$ & Lorentzian line profile half width of the $i^{th}$ spectral line\\
    $\mathsf{V}_i$ & The Voigt profile of the $i^{th}$ spectral line \\
    $\bar{\bm{V}}$ & The dense $(N_{\rm lines} \times N_{\rm pix})$ matrix of all Voigt profiles stacked vertically \\
    $\hat{\bm{V}}$ & The sparse $(N_{\rm lines} \times N_{\rm sparse})$ matrix of all Voigt profiles stacked vertically \\
    \hline
    \multicolumn{2}{c}{Scalars}\\
    \hline
    $N_{\rm lines}$ & Number of spectral lines \\
    $N_{\rm pix}$ & Number of pixel coordinates in the precomputed spectrum \\
    $N_{\rm sparse}$ & Number of non-zero pixels computed in the sparse implementation\\
    $\pm \Delta \lambda_{\mathrm{buffer}}$ & Buffer exceeding the red and blue limits of the data spectrum\\
    $P_{\rm rom}$ & The prominence threshold of spectral lines to include in cloning \\
    $v\sin{i}$ & Stellar projected rotational broadening at inclination $i$ and equatorial velocity $v$\\
    $RV$ & Radial velocity of the star\\
    \enddata
\end{deluxetable}
